---
title: "Impact of Severe Weather on US Population Health and Economic Consequence"
author: "Robert Begg"
date: "April 28, 2016"
output: 
  html_document:
    keep_md: true
---
# Synopsis
This report uses the data from the National Oceanic and Atmospheric Administration (NOAA) Storm Database to addresses the impact of severe weather on the US population health and economic consequnence.  Two key questions are answered:

1.Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
2.Across the United States, which types of events have the greatest economic consequences?
# Loading and Processing the RaW Data

##Load libraries

```{r Load_Libraries, warning=FALSE,results='hide',message=FALSE}
library(utils)
library(R.utils)
library(ggplot2)
library(dplyr)
library(tidyr)
```


## Get source Data

The data was download from [Coursera project](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).  The data is a text file in CSV format, with missing data represented as empty fields.

[Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) about the National Weather Service Storm data and an [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf).

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.
```{r Download_Data_File}
if(!file.exists("repdata-data-StormData.csv.bz2")){
        download.file(
                "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                "repdata-data-StormData.csv.bz2" )    
}
if(!file.exists("repdata-data-StormData.csv")){
        bunzip2("repdata-data-StormData.csv.bz2")    
}
```

## Loading the Data

The CSV data was read into a data frame object.  
```{r Load_Data, cache=TRUE}
StormData <- read.csv("repdata-data-StormData.csv")

dim(StormData)
```
There was `r dim(StormData)[1]` observations of `r dim(StormData)[2]` variables.
The first 10 rows of the data:
```{r Head}
head(StormData)
```

## Subsetting the Data

We are only interested in observations with non-zero values for population health and economic variables.

```{r Subset_Data}
HealthData <- StormData %>%
        filter(FATALITIES !=0 | INJURIES !=0) %>%
        select(EVTYPE,FATALITIES,INJURIES)
dim(HealthData)

EconomicData <- StormData %>%
        filter(PROPDMG !=0 | CROPDMG !=0) %>%
        select(EVTYPE,PROPDMG,PROPDMGEXP,CROPDMG, CROPDMGEXP)
dim(EconomicData)
```


## Converting Damage Values

The original data uses 3 significant digits in the PROPDMG and CROPDMG variables, and use the PROPDMGEXP and CROPDMGEXP variables to define the exponents.  Although the documentation indicates the exponent columns should only contain K (thousands), M (Millions), B (Billions) the levels command below shows the data actually contains a number of values.
```{r Explore_Exponents}
levels(EconomicData$PROPDMGEXP) # Unique values used in Property Damage Exponent Variable
levels(EconomicData$CRPODMGEXP) # Unique values used in Crop Damage Exponent Variable
```

The following code was used to convert the damage variables to numeric values.  Values are intepreted as follows:

1. 0 - 8 are interpreted as character representation of the exponents
2. Lower and upper case values for K, M, B follow the documentation (1e+03, 1e+06, 1e+09 repsectively)
3. Lower and upper case H is interpreted as Hundred or 1e+02 
4. Empty "", Blank " ", "-","?" and "+" characters are treated as missing, and treated as zero in the analysis.
First we convert the damage estimates into numeric values.

```{r convert_Exponents}
# Create lookup table to convert exponents to consisten exponent format
lookup <- list( Factor = c(""," ","?","+","-","H","h","k","K","m","M","b","B",
                            "0","1","2","3","4","5","6","7","8","9"),
                Value = c(1,1,1,1,1,1e+02,1e+02,1e+03,1e+03,1e+06,1e+06,1e+09,1e+09,
                          0,1e+01,1e+02,1e+03,1e+04,1e+05,1e+06,1e+07,1e+08,1e+9))
print(lookup)        

# Function to convert exponents
myConvert <- function(old) {
        new <- array(dim = length(old))
        
        for (i in 1:length(old)) {
                new[i] <- lookup$Value[lookup$Factor == old[i]]
        }
        return(new)
}
# Modify the Factor values by converting to consistent exponent format
levels(EconomicData$PROPDMGEXP) <- myConvert(levels(EconomicData$PROPDMGEXP))
levels(EconomicData$CROPDMGEXP) <- myConvert(levels(EconomicData$CROPDMGEXP))

EconomicData <- mutate(EconomicData, 
                       pDMG = PROPDMG * as.numeric(levels(PROPDMGEXP))[PROPDMGEXP] / 1e+09,
                       cDMG = CROPDMG * as.numeric(levels(CROPDMGEXP))[CROPDMGEXP] / 1e+09,
                       tDMG = pDMG + cDMG )

```


# Results

First we aggregate all fatalities and injuries by EVTYPE. 

```{r Aggregate_Health}
# Aggregate number of fatalities and injuries by event type, order result
healthType <- aggregate(cbind(FATALITIES, INJURIES) ~ EVTYPE, StormData, sum)
```

Then we remove observations that have zero fatalities or injuries.
```{r Subset_Health}
# Remove observations with no fatalitites or injuries
fType <- subset(healthType, FATALITIES!=0, select = c(EVTYPE,FATALITIES))

fType <- cbind(fType, round(fType$FATALITIES/sum(fType$FATALITIES)*100,2))
colnames(fType)[3] <- c("Percentage")

iType <- subset(healthType, INJURIES!=0, select = c(EVTYPE,INJURIES))

iType <- cbind(iType, round(iType$INJURIES/sum(iType$INJURIES)*100,2))
colnames(iType)[3] <- c("Percentage")
```

Finally we create a set ordered by Fatalities, and another ordered by Injuries.
```{r OrderHealh}
# Order result
fType <- fType[order(-fType$FATALITIES),]
iType <- iType[order(-iType$INJURIES),]
```

## Which Events are most harmful with respect to population health?

The data has two variables, fatalities and injuries, that indicate harm to population health.

```{r set_Top_Health_Events}
#set number of top events
n <- 35
```

### Fatalities

The following plot shows the Top `r n` Weather events (EVTYPE) causing fatalities:
```{r Plot_Fatalities, fig.width=8,fig.height=10}

fPlot <- ggplot(fType[1:n,], aes(EVTYPE, FATALITIES)) + 
        geom_bar(stat="identity", fill="firebrick",colour="yellow") +
        scale_x_discrete(limits=(fType[1:n,]$EVTYPE)) +
        ggtitle("Fatalities by Weather Event Type") +
        ylab("Number of Fatalities") +
        xlab("Weather Event Type") +
        theme(axis.title.x = element_text(face="bold", colour="black", size=12),
           axis.text.x  = element_text(angle=90, vjust=0.5, size=12)) +
        theme(axis.title.y = element_text(face="bold", colour="black", size=14),
              axis.text.y  = element_text(size=12)) +
        theme(title = element_text(face="bold", colour="black", size=16))
           
fPlot
```

### Injuries

The following plot shows the Top `r n` Weather events (EVTYPE) causing inuries:

```{r Plot_Injuries, fig.width=8,fig.height=10}

iPlot <- ggplot(iType[1:n,], aes(EVTYPE, INJURIES)) + 
        geom_bar(stat="identity", fill="navyblue",colour="yellow") +
        scale_x_discrete(limits=(iType[1:n,]$EVTYPE)) +
        ggtitle("Injuries by Weather Event Type") +
        ylab("Number of Injuries") +
        xlab("Weather Event Type") +
        theme(axis.title.x = element_text(face="bold", colour="black", size=12),
           axis.text.x  = element_text(angle=90, vjust=0.5, size=12)) +
        theme(axis.title.y = element_text(face="bold", colour="black", size=14),
              axis.text.y  = element_text(size=12)) +
        theme(title = element_text(face="bold", colour="black", size=16))
           
iPlot
```

### Conclusion

The events that cause 1 % or more of Fatalities:
```{r}
fType[fType$Percentage>=1,]$EVTYPE
```
The events that cause 1 % or more of Inuries:
```{r}
iType[iType$Percentage>=1,]$EVTYPE
```

## Which types of events have the greatest economic consequences?

The data has two variables, property damage and crop damage , that indicate economic consequence.

```{r set_Top_Damage_Events}
#set number of top events
n <- 20
```

Next we aggregate all damage by EVTYPE. 

```{r}
# Aggregate number of fatalities and injuries by event type, order result

dmgType <- aggregate(cbind(pDMG, cDMG) ~ EVTYPE, EconomicData, sum) %>%
        gather(dType, Damage, pDMG, cDMG) %>%
        arrange(desc(Damage))

# Order the Weather Events in order of damage, remove duplicates
orderedEcoEvents <- aggregate(cbind(tDMG) ~ EVTYPE, EconomicData, sum) %>%
        arrange(desc(tDMG))
        
# Refacttor EVTYPE to order by total damage
dmgType$EVTYPE <- factor(dmgType$EVTYPE, levels = orderedEcoEvents$EVTYPE)
                
```

Finally we create a set ordered by Fatalities, and another ordered by Injuries.
```{r OrderEconomic}
# Order result
pType <- dmgType[order(-dmgType$pDMG),]
cType <- dmgType[order(-dmgType$cDMG),]
```

### Property Damage

The following plot shows the Top `r n` Weather events (EVTYPE) causing property damage:
```{r Plot_Property_Damage, fig.width=8,fig.height=10}

pPlot <- ggplot(pType[1:n,], aes(EVTYPE, pDMG)) + 
        geom_bar(stat="identity", fill="firebrick",colour="yellow") +
        scale_x_discrete(limits=(pType[1:n,]$EVTYPE)) +
        ggtitle("Property Damage by Weather Event Type") +
        ylab("Property Damage (Billions $") +
        xlab("Weather Event Type") +
        theme(axis.title.x = element_text(face="bold", colour="black", size=12),
           axis.text.x  = element_text(angle=90, vjust=0.5, size=12)) +
        theme(axis.title.y = element_text(face="bold", colour="black", size=14),
              axis.text.y  = element_text(size=12)) +
        theme(title = element_text(face="bold", colour="black", size=16))
           
pPlot
```

### Crop Damage

The following plot shows the Top `r n` Weather events (EVTYPE) causing crop damage:
```{r Plot_Crop_Damage, fig.width=8,fig.height=10}

pPlot <- ggplot(cType[1:n,], aes(EVTYPE, cDMG)) + 
        geom_bar(stat="identity", fill="firebrick",colour="yellow") +
        scale_x_discrete(limits=(cType[1:n,]$EVTYPE)) +
        ggtitle("Crop Damage by Weather Event Type") +
        ylab("Crop Damage (Billions $)") +
        xlab("Weather Event Type") +
        theme(axis.title.x = element_text(face="bold", colour="black", size=12),
           axis.text.x  = element_text(angle=90, vjust=0.5, size=12)) +
        theme(axis.title.y = element_text(face="bold", colour="black", size=14),
              axis.text.y  = element_text(size=12)) +
        theme(title = element_text(face="bold", colour="black", size=16))
           
pPlot
```

### Total Damage

The following plot shows the Top `r n` Weather events (EVTYPE) causing crop damage:
```{r Plot_Total_Damage, fig.width=8,fig.height=10}

p <- ggplot(topEcoEvents, aes(x=EVTYPE, y=Damage, fill=dType)) + geom_bar(stat="identity")
p
```

